Using Optimal Time Step Selection to Boost the Accuracy of Finite Difference Methods for a Class of Time Dependent Partial Differential Equations

### 2008-05-13 papers/OTSPDE/optimal_time_steps_for_pdes.tex
OTSPDE:  first completed draft of paper.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First draft: 2008/05/??, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
In this article, we present a novel technique for obtaining high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  By using an optimally chosen the time step for a 
finite difference scheme (and possibly adding a few correction terms), we 
are able to boost the order of accuracy of low-order finite difference 
schemes.  Through simple numerical analysis arguments, we explain the origin 
of the boost in accuracy and calculate the computational cost of the
high-order methods obtained via optimal time step (OTS) selection.  We 
demonstrate the utility of OTS selection on a wide array of classical PDEs
in one and more space dimensions.   We find that OTS selection is applicable 
to linear and semilinear PDEs and that it is easily extended to problems 
on irregular domains. 
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{KEVIN T. CHU}
         {HIGH-ORDER ACCURATE FD SCHEMES VIA OPTIMAL TIME STEP SELECTION}

@@

High-order accurate methods that are simple combinations or modifications of 
low-order schemes abound.  A few well-known examples include Richardson 
extrapolation~\cite{richardson_1927,heath_book}, the modified nine-point 
formula for the Poisson equation~\cite{iserles_book}, and the Lax-Wendroff 
method for the advection equation~\cite{leveque_book_1992, leveque_book_2002,
gko_book}.  Cancellation of higher-order errors (whether fortuitous or 
intentional) is a key to principle underlying the superior accuracy of these 
each of methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for developing numerical methods for 
PDEs has its roots in the classical numerical analysis of PDEs, which tends to 
individually analyze the discretization of the terms in the PDE.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for \emph{all} 
independent variables.  Unfortunately, this common approach for developing 
and analyzing numerical methods is often suboptimal because it neglects the 
potential for achieving high order of accuracy using only simple finite 
difference schemes based on compact, low-order stencils.

Perhaps the greatest weakness of the conventional method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides all 
of the details about higher-order errors.  The information lost in the ``big-O" 
notation eliminates the possibility of reducing numerical error by taking 
advantage of relationships between the terms in the error.  The modified 
nine-point formula for the Poisson equation (also known under the name 
Mehrstellen)~\cite{iserles_book} and the Lax-Wendroff scheme for the 
advection equation~\cite{leveque_book_1992, leveque_book_2002, gko_book} 
demonstrate the power of explicitly retaining higher-order terms in the 
discretization error when analyzing numerical schemes.  For both of these 
methods, examination of higher-order terms allows low-order methods to be 
transformed into high-order methods with only minor modifications to the 
numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the false impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006} Zhao showed 
that appropriately choosing the value of $\theta$ as a function of the time 
step, $\dt$, in the theta-method\footnote{In~\cite{zhao_2006} refers 
to $\theta$ as the weight factor and the theta-method as the single-step 
trapezoidal method.}~\cite{iserles_book, gko_book} can decrease the error 
of finite element solutions to the diffusion equation.

In this paper, we explore a novel technique based on optimal time step (OTS)
selection that can be used to construct high-order finite difference schemes 
for time dependent PDEs using only formally low-order stencils.  The essential 
idea is that a carefully chosen time step (and possibly the addition of
a few correction terms) can eliminate low-order terms in the discretization 
error, allowing a formally low-order method to deliver high-order accuracy.  
The optimal time step is calculated through a straightforward analysis of the 
discretization error that uses information from the PDE to relate the 
leading-order terms.  Optimal time step selection combines the well-known 
approach of using information provided by the PDE to reduce numerical error 
with the less-common notion that optimal choice of numerical parameters can 
boost the effectiveness of algorithmically simple and computationally 
low-cost methods.

@@

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The fact that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done on 
optimizing the parameters of numerical methods (\eg time step and grid 
spacing) to improve accuracy.  To the best of the authors knowledge, Zhao's 
method for choosing an optimal value of $\theta$ in the theta-method is the 
only work that attempts to improve accuracy by modifying numerical parameters.
While he focuses solely on the optimal choice of $\theta$ for a fixed time 
step, his approach can be viewed as a form of optimal time step selection by 
reversing the roles of $\theta$ and the time step.

Before moving on, it is worth spending a few moments to draw the distinction
between optimal time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book,??} that are used in the context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection has the potential to produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

As a technique for constructing numerical methods, optimal time step selection 
has several desireable features.  Like any technique that leads to high-order 
methods, optimal time step selection produces schemes that greatly reduce the 
computational cost required to obtain a numerical solution of a desired 
accuracy.  However, its real power comes from the fact that it achieves
high-order accuracy while being based solely on very simple finite difference 
schemes.  For instance, we will see how optimal time step selection makes 
it possible to achieve 4th-order accuracy for the diffusion equation using only 
a second-order stencil for the Laplacian and forward Euler time integration.  
Moreover, high-order accuracy is not restricted to problems on simple, 
rectangular domains; irregular domains are straightforward to handle by
appropriately setting the values in ghost cells~\cite{fedkiw_1999, 
osher_fedkiw_book, gibou_2005}.  Optimal time step selection is even 
beneficial when using finite difference schemes to solve some nonlinear PDEs.  
While there are certainly limitations to optimal time step selection, we will 
see that there are a many problems where it is useful.

@@

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the FD 
approximation by optimizing the time step, there is no reason to explicitly
construct the scheme to be high-order.  In fact, using a high-order FD scheme 
only complicates the analysis required to calculate the optimal time 
step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the FD approximation, explicitly keeping track 
of the leading-order terms.  In contrast to conventional error analysis, we 
do not sweep all of the errors under ``big-O'' notation.  Rather, we take 
advantage of the fact that derivation of the leading-order terms in the 
error is straightforward for FD schemes through the use of Taylor series 
expansions.  Because each term in the error is proportional to a partial 
derivative of the solution, we can attempt to use the PDE to relate the 
leading-order terms to each other.  If we are fortunate (which we are
for many finite difference schemes, this careful analysis yields a direct 
relationship between the leading-order spatial and temporal errors that allows 
them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a polynomial in the numerical 
parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
$P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with the maximal order 
of accuracy for a given FD scheme.

An important class of FD schemes for time dependent PDEs are those based
on the method of lines.   When the method of lines approach is used for 
(\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
errors, and $\alpha$ and $\beta$ are positive constants that depend on the 
details of the PDE and FD scheme.  Setting $P = 0$, we find that the optimal 
time step is given by\footnote{The $\dt = 0$ root of $P$ is discarded 
because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
has a similar functional dependence on the grid spacing as a time step that 
satisfies a stability constraint.

Thus far, we have completely ignored the question of whether the FD scheme is 
stable when the time step is set to $\dto$.  Not surprisingly, OTS selection 
loses its value, when the optimal time step is greater than the smallest 
stable time step for an explicit scheme.  Fortunately, we can usually switch 
to an implicit (or semi-implicit) time integration scheme with the same order 
of accuracy for which the optimal time step is stable.  Note that the 
depending on how the original FD scheme is modified, a new optimal time step 
may need to be calculated.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the FD approximation of the spatial derivative 
operator acting on ${\mathbf u}$.  The second term in the local error arises 
directly from temporal discretization errors associated with the numerical 
scheme used to integrate the system of ODEs in time.  The first term arises 
because $F$ is only an approximation of the spatial derivative operators in 
the PDE.  As a result, the system of ODEs~(\ref{eq:method_of_lines}) has an 
error due to the spatial discretization at \emph{every} point in time.  For 
a spatial discretization error of order $r$, the error in 
(\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the error 
accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
is solved numerically.  Because the method of lines is so frequently used to 
construct FD approximations for time dependent PDEs, we will focus solely on 
FD schemes of this type for the remainder of the paper to keep the discussion 
simple.  It is straightforward to extend the results we obtain to more general 
FD schemes.

To analyze the accuracy of the FD scheme for~(\ref{eq:model_PDE}) when the 
optimal time step is used (for a given grid spacing), we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading term in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been removed.  As for the leading-order 
error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

From the local error, we can compute the global error of the numerical 
solution over extended intervals of time by using the heuristic argument
that the global error is equal to the local discretization 
error divided by $\dt$.  Using heuristic argument, which is rigorously 
justified in many texts on numerical methods such as~\cite{gko_book}, we 
find that the global error of the FD scheme using an optimal time step is 
\beq
O(\dx^p) + O(\dt^q).
\label{eq:global_error_ots}
\eeq
Observe that the global error for the FD scheme when an arbitrary time
step is used has exactly the same form as~(\ref{eq:global_error_ots}) with
$p$ and $q$ replaced by $r$ and $s$, respectively.  Thus, we see that 
using an optimal time step boosts the accuracy of the FD scheme.

It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
in order to derive the error estimate.  In other words, there is really only 
one parameter, $\dx$, that controls the numerical accuracy of the FD scheme.  
This important point will be discuss in further detail in the next section.

\subsubsection*{Formal vs. Practical Accuracy}
For time dependent PDEs, evaluating the order of accuracy for a numerical
method is subtle because there are formally two separate orders of accuracy 
to consider\footnote{The situation is more straightforward for 
time-independent PDEs because there is usually only one order of accuracy to 
be concerned with.}:  temporal and spatial.  While it is theoretically
interesting and important to understand how the error depends on both the 
grid spacing $\dx$ and the time step $\dt$, in practice the accuracy 
of a numerical scheme is always controlled by only one of the two numerical
parameters.

Even though spatial and temporal orders of accuracy for a numerical method
are formally separate, one will alway dominate for any given choice of 
$\dx$ and $\dt$.  For example, when solving the diffusion  equation 
using the backward Euler method for time integration with a second-order 
central difference stencil for the Laplacian, formal analysis shows that 
the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
stability constraints on the numerical parameters, the time
step and grid spacing are free to vary independently.  In this situation, the 
practical error for the method depends on the relative sizes of the time step 
and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
$O(\dt)$ which means that the error in the numerical solution is 
primarily controlled by the time step.  Similarly, when 
$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
the error is controlled by the grid spacing.  Finally, when 
$\dt  = O(\dx^2)$, the practical error is 
$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
primarily controlled by one of the two numerical parameters, and varying
the subdominant parameter while holding the dominant parameter fixed does 
not significantly affect the error.

When there are constraints on the numerical parameters, there is often 
less freedom in choosing the controlling parameter.  For instance, if we 
solve the diffusion equation using a forward Euler time integration scheme 
with a second-order central difference stencil for the Laplacian, stability
considerations require that we choose $\dt = O(\dx^2)$.  
Combining this stability constraint with the formal error for the scheme
shows that the practical error is 
$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
method is completely controlled by the grid spacing; the temporal error can 
never dominate the spatial error.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Manuscript started: 2007/09/01
% First completed draft: 2008/05/12, Vitamin D, Inc.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{abstract}
In this article, we present a novel technique for obtaining high-order
accurate numerical solutions to time dependent partial differential 
equations using only formally low-order finite difference stencils and
time integration schemes.  We achieve high-order accuracy in a simple
manner by using an optimally chosen time step (and possibly adding a few 
correction terms to the finite difference scheme).  Through straightforward 
numerical analysis arguments, we explain the origin of the boost in accuracy 
and estimate the computational cost of the high-order methods obtained via 
optimal time step (OTS) selection.  We demonstrate the utility of OTS 
selection on a wide array of classical PDEs in one and more space dimensions
and find that it is applicable to linear and semilinear PDEs on both regular 
and irregular domains. 
\end{abstract}

@@

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{KEVIN T. CHU}
         {HIGH-ORDER FINITE DIFF SCHEMES VIA OPTIMAL TIME STEP SELECTION}

@@

There are many high-order accurate methods that are simple combinations or 
modifications of low-order schemes.  A few well-known examples include 
Richardson extrapolation~\cite{richardson_1927, heath_book}, the modified 
nine-point formula for the 2D Poisson equation~\cite{iserles_book}, and the 
Lax-Wendroff method for the advection equation~\cite{leveque_book_1992, 
leveque_book_2002, gko_book}.  Cancellation of higher-order errors (whether 
fortuitous or intentional) is a key principle underlying the superior 
accuracy of these methods.

More commonly, high-order accurate numerical methods are constructed by 
requiring that the discretization error possesses the desired formal order of 
accuracy for \emph{all} of the independent variables in the PDE.  While this 
constraint is certainly sufficient for high-order accuracy, it is by no means 
necessary.  This conventional approach for constructing numerical methods 
has its roots in the classical numerical analysis of PDEs, which tends to 
analyze the discretization of the terms in the PDE individually.  Within this 
framework, it is natural to seek high-order accuracy by ensuring that the 
numerical scheme is simultaneously high-order accurate for all independent 
variables.  Unfortunately, this common approach for developing and analyzing 
numerical methods is often suboptimal because it neglects the potential for 
achieving high-order accuracy using only simple, low-order schemes.

Perhaps the greatest weakness of the standard method for constructing 
high-order numerical methods is the use of ``big-O" notation, which hides 
the details of the discretization error.  The information lost in the 
``big-O" notation eliminates the possibility of reducing numerical error by 
taking advantage of relationships between the terms in the error.  The 
modified nine-point formula for the Poisson equation~\cite{iserles_book} 
and the Lax-Wendroff scheme for the advection equation~\cite{leveque_book_1992,
leveque_book_2002, gko_book} demonstrate the power of explicitly retaining 
higher-order terms in the discretization error when analyzing numerical 
schemes.  For both of these methods, examination of higher-order terms allows 
low-order methods to be transformed into higher order methods with only minor 
modifications to the numerical scheme.

The usual approach for developing numerical methods has another significant
shortcoming -- it places no value on the optimization of numerical parameters 
(\eg $\dx$ and $\dt$) for accuracy.  Again, ``big-O" notation is 
somewhat to blame for this situation because it gives the false impression 
that the error depends on each numerical parameter independently.  Aside
from constraints needed for numerical stability, ``big-O" notation makes 
it appear as though any combination of numerical parameters is equally good 
as long as they are all ``small enough.''  For some problems, however, 
optimizing the numerical parameters can significantly increase the accuracy
of the numerical solution.  For example, in~\cite{zhao_2006}, Zhao showed 
that choosing an appropriate value for $\theta$ as a function of the time 
step in the theta-method\footnote{In~\cite{zhao_2006} refers to $\theta$ as 
the weight factor and the theta-method as the single-step trapezoidal 
method.}~\cite{iserles_book, gko_book} can decrease the error of finite 
element solutions to the diffusion equation.

In this paper, we explore a novel technique for constructing high-order 
finite difference methods for time dependent PDEs using only formally 
low-order stencils and time integration schemes.  The basic idea is that a 
carefully chosen time step (and possibly the addition of a few correction 
terms) can eliminate low-order terms in the discretization error, allowing a 
formally low-order method to deliver high-order accuracy.  The optimal time 
step is calculated in a straightforward manner by using knowledge of the 
PDE to eliminate the leading-order terms in the discretization error.  
Optimal time step selection combines the well-known approach of using 
the PDE to reduce numerical error with the less-common notion that optimal 
choice of numerical parameters can boost the effectiveness of algorithmically 
simple and computationally low-cost methods.

@@

Optimal time step selection as a technique for improving the overall accuracy 
of numerical schemes for time dependent PDEs appears to be a novel idea.
Typically, the primary concern when choosing a time step is numerical 
stability with accuracy being a secondary consideration.  The fact that a 
clever choice of time step can be used to dramatically reduce the error is 
rarely considered.  In general, little research seems to have been done in
the area of optimizing the parameters of numerical methods (\eg time step and 
grid spacing) to improve accuracy.  To the best of the author's knowledge, 
Zhao's method for choosing an optimal value of $\theta$ in the theta-method is 
the only work that attempts to improve accuracy by modifying numerical 
parameters.  While he focuses solely on the optimal choice of $\theta$ for a 
fixed time step, his approach can be viewed as a form of optimal time step 
selection by reversing the roles of $\theta$ and the time step.

It is worth spending a few moments to draw the distinction between optimal 
time step selection and traditional adaptive time stepping 
techniques~\cite{iserles_book,??} that are used in the context of the method 
of lines~\cite{iserles_book, gko_book}.  On the surface, optimal time step 
selection may appear to be just a crude version of adaptive time step control.
However, the two methods are actually very different.  While both methods seek 
to reduce numerical error by controlling the time step, the numerical errors 
they control are fundamentally different.  Adaptive time stepping can only 
be used to reduce temporal discretization errors because it controls the 
errors that arise when solving the coupled system of ODEs for the 
semi-discrete approximation to the PDE; the spatial discretization errors 
are completely controlled by the finite difference scheme selected to 
approximate the spatial derivatives.  
In contrast, optimal time step selection simultaneously reduces 
\emph{both} the spatial and temporal discretization errors because it
uses information from the PDE to choose the time step.  Thus, optimal time 
step selection can potentially produce superior results compared with 
traditional adaptive time stepping approaches for integrating the 
semi-discrete equations associated with time dependent PDEs.  Another 
important distinction between the two techniques is that optimal time step 
selection uses a \emph{fixed} time step which makes it significantly less 
computationally complex than adaptive time stepping methods.

As a technique for constructing numerical methods, optimal time step selection 
has several desirable features.  Like any technique that leads to high-order 
methods, optimal time step selection produces schemes that greatly reduce the 
computational cost required to obtain a numerical solution of a desired 
accuracy.  However, its real power comes from the fact that it achieves
high-order accuracy while being based solely on very simple finite difference 
schemes.  For instance, we will see how optimal time step selection makes 
it possible to achieve 4th-order accuracy for the diffusion equation using only 
a second-order stencil for the Laplacian and forward Euler time integration.  
Moreover, high-order accuracy is not restricted to problems on simple, 
rectangular domains; irregular domains are straightforward to handle by
appropriately setting the values in ghost cells~\cite{gibou_2005, fedkiw_1999, 
osher_fedkiw_book}.  Optimal time step selection is even beneficial when 
using finite difference schemes to solve some nonlinear PDEs.  While there 
are certainly limitations to optimal time step selection, we will see that 
there are many problems where it is useful.

@@

\section{\label{sec:optimal_time_step_selection} 
         Optimal Time Step (OTS) Selection}
Optimal time step selection is not by itself a method for constructing 
finite difference schemes.  Rather it is a way to enhance to the performance
of existing finite difference schemes by carefully choosing the time step to 
eliminate low-order numerical errors.  The two basic ideas underlying 
optimal time step selection are (1) the PDE provides valuable insight into the 
discretization errors for finite difference schemes and (2) a judicious choice 
of time step can be used to eliminate the leading-order (and possibly 
higher-order) terms in the error.  In this section we present and analyze the 
technique of optimal time step selection for finite difference schemes 
constructed to solve evolution equations of the form 
\beq
  \frac{\partial u}{\partial t} = 
    F\left(x, u(x), D u(x), D^2 u(x), \ldots \right), 
\eeq
where the right hand side is an arbitrary function of the solution $u$, its 
spatial derivatives, and the independent variable $x$.
Because this class of time dependent PDEs is so prevalent and important in 
many problems of applied mathematics, restricting our attention to problems 
of this form is not a serious limitation.

\subsection{\label{sec:ots_model_pde} 
            Optimal Time Step for a Model PDE}
Let us begin by considering a finite difference approximation for the
following time dependent PDE in one spatial dimension: 
\beq
  \frac{\partial u}{\partial t} = A \frac{\partial^n u}{\partial x^n},
  \label{eq:model_PDE}
\eeq
where $u$ is the solution to the PDE, $n$ is the spatial order of the 
PDE, and $A$ is a constant of the appropriate sign so that~(\ref{eq:model_PDE}) 
is well-posed.  Because we will be boosting the accuracy of the finite 
difference approximation by optimizing the time step, there is no reason to 
explicitly construct the scheme to be high-order.  In fact, using a high-order 
finite difference scheme only complicates the analysis required to calculate 
the optimal time step.

To compute the optimal time step, we first analyze the discretization errors 
(both spatial and temporal) for the finite difference approximation, 
explicitly keeping track of the leading-order terms.  In contrast to 
conventional error analysis, we do not sweep all of the errors under ``big-O'' 
notation.  Rather, we take advantage of the fact that derivation of the 
leading-order terms in the error is straightforward for finite difference
schemes through the use of Taylor series expansions.  Because each term in 
the error is proportional to a partial derivative of the solution, we can 
attempt to use the PDE to relate the leading-order terms to each other.  If we 
are fortunate (which we are for many finite difference schemes), this careful 
analysis yields a direct relationship between the leading-order spatial and 
temporal errors that allows them to be combined as a single term of the form:
\beq
  (L u) P(\dx, \dt) ,
  \label{eq:leading_order_error_model_PDE_general}
\eeq
where $L$ is a differential operator and $P$ is a polynomial in the numerical 
parameters $\dx$ and $\dt$.  By selecting the numerical parameters so that 
$P = 0$, we can completely eliminate the leading-order term in the 
discretization error.  Because time dependent PDEs are often solved by 
stepping in time, it is natural to choose the time step to be a function of 
the grid spacing.  The time step selected in this way is the
\emph{optimal time step} for the finite difference scheme.  Using the optimal
time step for time stepping yields a numerical solution with a boosted order 
of accuracy for a given finite difference scheme.

An important class of finite difference schemes for time dependent PDEs are 
those based on the method of lines.   When the method of lines approach is 
used for (\ref{eq:model_PDE}), $P$ takes a particularly simple 
form:
\beq
  P(\dx, \dt) = (\alpha \dx^r - \beta \dt^s) \dt,
  \label{eq:leading_order_error_model_PDE_MOL}
\eeq
where $r$ and $(s+1)$ are the orders of the leading spatial and temporal 
discretization errors, and $\alpha$ and $\beta$ are positive constants that 
depend on the details of the PDE and finite difference scheme.  Setting 
$P = 0$, we find that the optimal time step is given by\footnote{The $\dt = 0$ 
root of $P$ is discarded because it is not numerically meaningful.}
\beq
  \dto = \gamma \dx^{r/s},
  \label{eq:optimal_time_step}
\eeq
where $\gamma = (\alpha/\beta)^{1/s}$.  Notice that the optimal time step 
has a similar functional dependence on the grid spacing as a time step that 
satisfies a stability constraint.

\subsubsection*{\label{sec:error_analysis} 
            Error Analysis}
The form of the local error in 
(\ref{eq:leading_order_error_model_PDE_MOL}) follows from the fact 
that the method of lines first approximates the PDE as a system of ODEs in 
time by only discretizing the spatial derivatives: 
\beq
{\mathbf u}_t = F({\mathbf u}),
\label{eq:method_of_lines}
\eeq
where ${\mathbf u}$ is the vector of values of $u$ at the grid points and
$F({\mathbf u})$ is the finite difference approximation of the spatial 
derivative operator acting on ${\mathbf u}$.  The second term in the local 
error arises directly from temporal discretization errors associated with the 
numerical scheme used to integrate the system of ODEs in time.  The first 
term arises because $F$ is only an approximation of the spatial derivative 
operators in the PDE.  As a result, the system of 
ODEs~(\ref{eq:method_of_lines}) has an error due to the spatial discretization 
at \emph{every} point in time.  For a spatial discretization error of order 
$r$, the error in (\ref{eq:method_of_lines}) is $O(\dx^r)$.  Therefore, the 
error accumulated during each time step due to spatial discretization error is 
$O(\dt \dx^r)$ regardless of how the system of ODEs (\ref{eq:method_of_lines})
is solved numerically.  Because the method of lines is so frequently used to 
construct finite difference approximations for time dependent PDEs, we will 
focus solely on finite difference schemes of this type for the remainder of 
the paper to keep the discussion simple.  It is straightforward to extend the 
results we obtain to more general finite difference schemes.

To analyze the accuracy of the finite difference scheme 
for~(\ref{eq:model_PDE}) when the optimal time step is used, we examine the 
higher-order terms in the discretization error following the conventional 
approach (\ie ``big-O'' notation).  Elimination of the leading terms in the 
discretization error leaves a local error of the form 
$O(\dt \dx^p) + O(\dt^{q+1})$, where $p>r$ and $(q+1) > (s+1)$ are the orders 
of the spatial and temporal discretization errors, respectively, after the 
leading-order terms in the error have been removed.  As for the leading-order 
error (\ref{eq:leading_order_error_model_PDE_MOL}), this form of the local 
error is a direct consequence of the method of lines approach to discretizing 
the PDE.

From the local error, we can compute the global error of the numerical 
solution over extended intervals of time by using the heuristic argument
that the global error is equal to the local error divided by $\dt$.  Using 
this heuristic argument, which is rigorously justified in many texts on 
numerical methods such as~\cite{gko_book}, we find that the global error of 
the finite difference scheme using an optimal time step is 
\beq
O(\dx^p) + O(\dt^q).
\label{eq:global_error_ots}
\eeq
Observe that the global error for the finite difference scheme when an 
arbitrary time step is used has exactly the same form 
as~(\ref{eq:global_error_ots}) with $p$ and $q$ replaced by $r$ and $s$, 
respectively.  Thus, because $p > r$ and $q > s$, we see that using an optimal 
time step boosts the overall accuracy of the finite difference scheme.

It is important to emphasize that $\dx$ and $\dt$ are \emph{not} independent 
in~(\ref{eq:global_error_ots}) because the time step must be set to $\dto$ 
in order to derive the error estimate.  In other words, there is really only 
one parameter, $\dx$, that controls the numerical accuracy of the finite 
difference scheme.  This important point will be discuss in further detail in 
the following section.

\subsubsection*{Formal vs. Practical Accuracy}
For time dependent PDEs, evaluating the order of accuracy for a numerical
method is subtle because there are formally two separate orders of accuracy 
to consider\footnote{The situation is more straightforward for 
time-independent PDEs because there is usually only one order of accuracy to 
be concerned with.}:  temporal and spatial.  While it is theoretically
interesting and important to understand how the error depends on both the 
grid spacing $\dx$ and the time step $\dt$, in practice, the accuracy 
of a numerical scheme is always controlled by only one of the two numerical
parameters.

Even though spatial and temporal orders of accuracy for a numerical method
are formally separate, one will always dominate for a given choice of 
$\dx$ and $\dt$.  For example, when solving the 1D diffusion equation 
using the backward Euler method for time integration with a second-order 
central difference stencil for the Laplacian, formal analysis shows that 
the global error is $O(\dx^2) + O(\dt)$.  Since there are no 
stability constraints on the numerical parameters, the time
step and grid spacing are free to vary independently.  In this situation, the 
practical error for the method depends on the relative sizes of the time step 
and grid spacing.  When $\dt \gg \dx^2$, the practical error is 
$O(\dt)$ which means that the error in the numerical solution is 
primarily controlled by the time step.  Similarly, when 
$\dt \ll \dx^2$, the practical error is $O(\dx^2)$ so that 
the error is controlled by the grid spacing.  Finally, when 
$\dt  = O(\dx^2)$, the practical error is 
$O(\dt) = O(\dx^2)$.  In all cases, the practical error is 
primarily controlled by one of the two numerical parameters, and varying
the subdominant parameter while holding the dominant parameter fixed does 
not significantly affect the error.

When there are constraints on the numerical parameters, there is less freedom 
in choosing the controlling parameter.  For instance, if we solve the 
1D diffusion equation using a forward Euler time integration scheme with a 
second-order central difference stencil for the Laplacian, stability
considerations require that we choose $\dt = O(\dx^2)$.  
Combining this stability constraint with the formal error for the scheme
shows that the practical error is 
$O(\dx^2) + O(\dt) = O(\dx^2)$.  Thus, the accuracy of the 
method is completely controlled by the grid spacing; the temporal error can 
never dominate the spatial error.

Numerical methods for time dependent PDEs are commonly designed so that 
the global error is primarily controlled by the grid spacing.  The numerical
errors introduced by the time integration scheme are typically \emph{forced}
to be subdominant to the spatial discretization errors manually or as a result
of stability constraints.  In other words, the operational procedure used 
when developing numerical methods is usually to first choose a spatial 
discretization scheme and grid spacing so that a desired level of accuracy 
is achieved and then select the time integration scheme and time step in 
such a manner that the temporal errors do not overcome the spatial errors.  
This methodology explains the tendency to evaluate numerical schemes for
time dependent PDEs predominantly on their order of accuracy with respect 
to the grid spacing.
